Synthesizing Quotient Lenses

### 2018-03-15 papers/QuotientLenses/main.tex
slight reworking of the first couple paragraphs

Building on recent work that showed how to synthesize bijective lenses from high
level specifications, we offer three innovations to synthesize quotient lenses.
First, we introduce {\em quotient regular expressions} (QREs), that easily mark
aspects of data formats that are inessential and hence give a high-level
specification of regular languages with an equivalence relation defined on
them. Next, we introduce {\em QRE lenses}, i.e., lenses mapping between QREs.
Our key technical result is a proof that every QRE lens can be transformed into
one of the form $c ; \ell ; d^{-1}$, where $c$ and $d^{-1}$ are {\em
canonizers}, $\ell$ is a bijective transformation, and $;$ is
composition\iflastminute\bcp{$c_2^{-1}$ is ugly, but I think we need the
superscript for accuracy.  Perhaps calling them $c$ and $d$ instead of $c_1$
and $c_2$ would be better?}\fi.  Intuitively, the canonizers normalize data by
picking a canonical representative of the relevant equivalence relations. This
theorem means we can push all normalization to the outer edges of a
bidirectional transformation, leaving in the middle a pure bijective
transformation. Finally, we leverage this normalization theorem to {\em
synthesize} QRE lenses from a pair of QREs and example input-output pairs,
reusing earlier work on synthesizing plain bijective lenses. We have
implemented QREs and QRE lens synthesis in the bidirectional programming
language Boomerang. We evaluate the effectiveness of our approach by using it
to synthesize QRE lenses between various real-world data formats in the
Optician benchmark suite.

@@

\section{Introduction}
Programmers often need to write programs that bidirectionally convert data
between two different formats.\bcp{It's a shame to lose the concrete
  examples here.} A time-tested and appealing way to do this is
to use \emph{lenses}~\cite{Lenses}, programs that simultaneously define pairs of
functions for translating the data from the source format to the target format
(the \emph{get} direction) and back again (the \emph{put} direction). The
main reason that lenses are appealing is because lens programming languages
guarantee that the \kw{put} and \kw{get} functions of a lens satisfy
certain \emph{lens laws} constraining how the two functions interact with each
other.

Recent work~\cite{optician} developed a tool called Optician to synthesize a
subset of the {\em bijective lenses} that can be derived in the lens
programming language {\em Boomerang}~\cite{boomerang}. Each lens $\ell : S
\Leftrightarrow T$ that Optician can synthesize specifies a bijection from the
language of a source regular expression $S$ to a target language of a target
regular expression $T$: \begin{equation}\label{bijectivelenslaws} \ell.\get \;
(\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s) = s
\end{equation}
Optician makes programming bijective lenses faster and easier, and though
many useful lenses are not bijective in nature, they do have a bijective core
surrounded by structured transformations. Hence, Optician can be used as a
component in a system that synthesizes more complex lenses.

>>>>

Building on recent work that showed how to synthesize bijective lenses from high
level specifications, we offer three innovations to synthesize quotient lenses.
First, we introduce {\em quotient regular expressions} (QREs), that easily mark
aspects of data formats that are inessential and hence give a high-level
specification of regular languages with an equivalence relation defined on
them. Next, we introduce {\em QRE lenses}, i.e., lenses mapping between QREs.
Our key technical result is a proof that every QRE lens can be transformed into
one of the form $c ; \ell ; d^{-1}$, where $c$ and $d^{-1}$ are {\em
canonizers}, $\ell$ is a bijective transformation, and $;$ is
composition.  Intuitively, the canonizers normalize data by
picking a canonical representative of the relevant equivalence relations. This
theorem means we can push all normalization to the outer edges of a
bidirectional transformation, leaving in the middle a pure bijective
transformation. Finally, we leverage this normalization theorem to {\em
synthesize} QRE lenses from a pair of QREs and example input-output pairs,
reusing earlier work on synthesizing plain bijective lenses. We have
implemented QREs and QRE lens synthesis in the bidirectional programming
language Boomerang. We evaluate the effectiveness of our approach by using it
to synthesize QRE lenses between various real-world data formats in the
Optician benchmark suite.

@@

\section{Introduction}
Programmers often need to write programs that bidirectionally convert data
between two different formats. For example, bibliographic data may be stored in
\bibtex{} or EndNote formats, spreadsheet data can be represented as either CSV
and TSV files, and web APIs can return data as either JSON or XML objects.  A
time-tested and appealing way to implement such conversions is to use
\emph{lenses}~\cite{Lenses}, programs that simultaneously define pairs of
functions for translating the data from the source format to the target format
(the \emph{get} direction) and back again (the \emph{put} direction).  Lens
programming languages guarantee that these \kw{put} and \kw{get} functions
satisfy certain \emph{lens laws}, which imply that round trips, from source to
target and back again, behave properly.

Recent work~\cite{optician} developed a tool called Optician to synthesize a
subset of the {\em bijective lenses} that can be derived in the lens
programming language {\em Boomerang}~\cite{boomerang}. Each lens $\ell : S
\Leftrightarrow T$ that Optician can synthesize specifies a bijection from the
language of a source regular expression $S$ to a target language of a target
regular expression $T$: \begin{equation}\label{bijectivelenslaws} \ell.\get \;
(\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s) = s
\end{equation}
Optician makes programming bijective lenses easier.  However, many useful lenses
are not strictly bijective in nature---the desired transformation might ignore
whitespace or the exact ordering of data fields, for instance---and it is a
shame that Optician's synthesis algorithm cannot be directly applied in such
cases.  One observation, however, is that non-bijective transformations can
often be structured as a bijective ``core'' surrounded by some kind of data
normalization at the edges. We can therefore hope to use the Optician algorithm
as component in a system that synthesizes more complex lenses.
